Approximation as Differentiation

### 2025-04-15 appendix/notes/grothendieck.tex
LatGal(X,Y) actually a semilattice.

\begin{proposition}
Suppose $C$ locally small. If $C$ has binary biproducts and set-indexed products, then $\Fam(C)$ is Cartesian
closed.
\end{proposition}

\begin{proof}
Suppose $C$ is locally small, with binary biproducts given by $(\biprod,\biinj,\biproj)$ and set-indexed
products given by $(\prod,\eval_{\prod},\lambda_{\prod})$. Write $(\times,\pi_1,\pi_2)$ for the product in
$\Fam(C)$ inherited from the biproduct in $C$. For any objects $X, Y, Z$ in $\Fam(C)$ define:

\begin{enumerate}
\item $[X, Y]$ to be the object $(X',Y')$ in $\Fam(C)$ with $X' = \Fam(C)(X,Y)$ and $Y'_f =
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}}$ for every $f: X \to Y$.
\item The family of isomorphisms $\lambda_{X,Y,Z}$ (natural in $X$) sending any $(f, g): X \times Y \to Z$ to
$(f', g'): X \to [Y, Z]$ with:
\begin{itemize}
\item $f'(x) = (f \comp (x, -), \reindex{g}{(x,-)} \comp \biinj_{\faml{Y}}):
Y \to Z$ where $\biinj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g}{(x,-)} \comp \biinj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f})$ where $\biinj_{\faml{X}_x}: \const_{\idx{Y}}(\faml{X}_x)
\to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\end{itemize}
\item The family of morphisms $\eval_{X,Y}: [X, Y] \times X \to Y$ in $\Fam(C)$ with
\begin{itemize}
\item $\idx{\eval_{X,Y}}(f, x) = \idx{f}(x)$
\item $\faml{\eval_{X,Y}}_{f, x} = \coprodM{{\eval_{\prod}}_x}{\faml{f}_x}:
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}} \biprod \faml{X}_x \to \faml{Y}_{\idx{f}(x)}$
\end{itemize}
\end{enumerate}

@@

To show $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f,g)$, suppose $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f',
g')$. Then:
\begin{itemize}
\item $f'(x) = (f^\dagger \comp (x, -), \reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{Y}}): Y \to Z$ where
$\biinj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f^\dagger})$ where $\biinj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$.
\end{itemize}
\noindent To show $f' = f : \idx{X} \to \Fam(C)(Y,Z)$, suppose $x \in \idx{X}$ and show $f'(x) = f(x): Y \to
Z$. Suppose $y \in \idx{Y}$.
\begin{enumerate}
\item $\idx{f'(x)}(y) = (f^\dagger \comp (x, -))(y) = f^\dagger(x,y) = \idx{f(x)}(y)$.
\item $\faml{f'(x)}_y = (\reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{Y}})_y = \reindex{g^\dagger}{(x,-)}_y \comp
{\biinj_{\faml{Y}}}_y = g^\dagger_{x,y} \comp {\biinj_{\faml{Y}}}_y = \coprodM{{\eval_{\prod}}_y \comp
g_x}{\faml{f(x)}_y} \comp {\biinj_{\faml{Y}}}_y = \faml{f(x)}_y$.
\end{enumerate}

@@

\item $\eval_{Y,Z} \comp (\lambda_{X,Y,Z}(f), \id_{Y}) = f$. Suppose $f: X \times Y \to Z$ in $\Fam(C)$
and $(x, y) \in \idx{X} \times \idx{Y}$. The diagram on the left below commutes by the universal property of
the set-indexed product; the commutativity of the diagram on the right is trivial.

@@

The diagram on the left below then commutes by Proposition~\ref{prop:biproduct:prod-coprod}, with
$\coprodM{\faml{f}_{x,y}\,\comp\,\biinj_{\faml{X}_x}}{\faml{f}_{x,y}\,\comp\,\biinj_{\faml{Y}_y}} =
\faml{f}_{(x,y)}$ by the uniqueness of the coproduct of morphisms. Then the diagram on the right in $\Fam(C)$
commutes by the definitions of $\lambda_{X,Y,Z}, \eval_{Y,Z}$ and $[Y,Z]$.

>>>>

\begin{proposition}
Suppose $\cat{C}$ locally small. If $\cat{C}$ has binary biproducts and set-indexed products, then
$\Fam(\cat{C})$ is Cartesian closed.
\end{proposition}

\begin{proof}
Suppose $\cat{C}$ is locally small, with binary biproducts given by $(\biprod,\biinj,\biproj)$ and set-indexed
products given by $(\prod,\eval_{\prod},\lambda_{\prod})$. Write $(\times,\pi_1,\pi_2)$ for the product in
$\Fam(\cat{C})$ inherited from the biproduct in $\cat{C}$. For any objects $X, Y, Z$ in $\Fam(\cat{C})$
define:

\begin{enumerate}
\item $[X, Y]$ to be the object $(X',Y')$ in $\Fam(\cat{C})$ with $X' = \Fam(\cat{C})(X,Y)$ and $Y'_f =
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}}$ for every $f: X \to Y$.
\item The family of isomorphisms $\lambda_{X,Y,Z}$ (natural in $X$) sending any $(f, g): X \times Y \to Z$ to
$(f', g'): X \to [Y, Z]$ with:
\begin{itemize}
\item $f'(x) = (f \comp (x, -), \reindex{g}{(x,-)} \comp \biinj_{\faml{Y}}):
Y \to Z$ where $\biinj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g}{(x,-)} \comp \biinj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f})$ where $\biinj_{\faml{X}_x}: \const_{\idx{Y}}(\faml{X}_x)
\to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\end{itemize}
\item The family of morphisms $\eval_{X,Y}: [X, Y] \times X \to Y$ in $\Fam(\cat{C})$ with
\begin{itemize}
\item $\idx{\eval_{X,Y}}(f, x) = \idx{f}(x)$
\item $\faml{\eval_{X,Y}}_{f, x} = \coprodM{{\eval_{\prod}}_x}{\faml{f}_x}:
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}} \biprod \faml{X}_x \to \faml{Y}_{\idx{f}(x)}$
\end{itemize}
\end{enumerate}

@@

To show $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f,g)$, suppose $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f',
g')$. Then:
\begin{itemize}
\item $f'(x) = (f^\dagger \comp (x, -), \reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{Y}}): Y \to Z$ where
$\biinj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f^\dagger})$ where $\biinj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$.
\end{itemize}
\noindent To show $f' = f : \idx{X} \to \Fam(\cat{C})(Y,Z)$, suppose $x \in \idx{X}$ and show $f'(x) = f(x): Y
\to Z$. Suppose $y \in \idx{Y}$.
\begin{enumerate}
\item $\idx{f'(x)}(y) = (f^\dagger \comp (x, -))(y) = f^\dagger(x,y) = \idx{f(x)}(y)$.
\item $\faml{f'(x)}_y = (\reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{Y}})_y = \reindex{g^\dagger}{(x,-)}_y \comp
{\biinj_{\faml{Y}}}_y = g^\dagger_{x,y} \comp {\biinj_{\faml{Y}}}_y = \coprodM{{\eval_{\prod}}_y \comp
g_x}{\faml{f(x)}_y} \comp {\biinj_{\faml{Y}}}_y = \faml{f(x)}_y$.
\end{enumerate}

@@

\item $\eval_{Y,Z} \comp (\lambda_{X,Y,Z}(f), \id_{Y}) = f$. Suppose $f: X \times Y \to Z$ in $\Fam(\cat{C})$
and $(x, y) \in \idx{X} \times \idx{Y}$. The diagram on the left below commutes by the universal property of
the set-indexed product; the commutativity of the diagram on the right is trivial.

@@

The diagram on the left below then commutes by Proposition~\ref{prop:biproduct:prod-coprod}, with
$\coprodM{\faml{f}_{x,y}\,\comp\,\biinj_{\faml{X}_x}}{\faml{f}_{x,y}\,\comp\,\biinj_{\faml{Y}_y}} =
\faml{f}_{(x,y)}$ by the uniqueness of the coproduct of morphisms. Then the diagram on the right in
$\Fam(\cat{C})$ commutes by the definitions of $\lambda_{X,Y,Z}, \eval_{Y,Z}$ and $[Y,Z]$.
